\Large{Adversarial Variational Optimization of Non-Differentiable Simulators}

### 2017-06-29 tex/paper.tex
save

Complex computer simulators are increasingly used across fields of science to
describe generative models tying parameters of an underlying theory to
experimental observations. Inference in this setup is often
difficult, as simulators rarely provide a way to directly evaluate the likelihood
function for a given observation. In this note, we develop a likelihood-free
inference algorithm for fitting a forward non-differentiable generative model to
observed data. We adapt the adversarial training procedure of generative
adversarial networks by replacing the implicit generative network with a
domain-based scientific simulator, and solve the resulting non-differentiable
minimax problem by minimizing variational upper bounds of the adversarial objectives.
Effectively, the procedure results in learning an arbitrarily tight
proposal distribution over simulator parameters, such that the corresponding
marginal distribution of the generated data matches the observations.
\glnote{Mention experimental results.}
\glnote{Add 'so what?' conclusion.}

@@

In many fields of science such as particle physics, climatology or population
genetics, computer simulators are used to describe complex processes that tie
parameters of an underlying theory to high dimensional observations. In most
cases, these implicit generative models~\citep{2016arXiv161003483M} are
specified as procedural implementations of forward stochastic processes that
generate data. Because it is usually computationally intractable, most simulators
do not provide a way to directly evaluate the likelihood function for a
given observation, thereby making inference difficult.
In addition, most scientific simulators are written using opaque low-level programming
languages, hence raising the bar for likelihood-free inference
algorithms relying e.g. on the simulator being a function with computable derivatives~\citep{2016arXiv160507826G}
or being a controllable probabilistic program~\citep{2016arXiv161009900L}.

In this note, we develop a likelihood-free inference algorithm for the point
estimation of the parameters of a forward non-differentiable
generative model. We adapt the adversarial
training procedure of generative adversarial
networks~\cite{goodfellow2014generative} by replacing the implicit generative
network with a domain-based scientific simulator, and solve the resulting
non-differentiable minimax problem by minimizing variational upper
bounds~\citep{2011arXiv1106.4487W,2012arXiv1212.4507S} of the adversarial
objectives. The procedure results in learning a proposal distribution over
simulator parameters, hence producing an arbitrarily tight family of models whose
joint collection of generated samples matches the observed data.
\glnote{add note on discrete data}

@@

Practically, the variational objectives \ref{eqn:vo-ud}-\ref{eqn:vo-ug}
have the effect of replacing the modeled data distribution of Eqn.~\ref{eqn:p_theta} with
a distribution parameterized in terms of $\psi$:
\begin{equation}\label{eqn:p_psi}
    \mathbf{x} \sim p(\mathbf{x}|\psi) \equiv \theta \sim q(\theta|\psi), \mathbf{z} \sim p(\mathbf{z}|\theta), \mathbf{x} = g(\mathbf{z}; \theta).
\end{equation}
Intuitively, this corresponds to a family of simulators, each configured
with randomly sampled parameters $\theta \sim q(\theta|\psi)$, whose joint collection
of generated samples is optimized with adversarial training to approach the real data distribution $p_r(\mathbf{x})$.
More formally, the learned model $p(\mathbf{x}|\psi)$ therefore corresponds to the marginal distribution
$\int q(\theta|\psi) p(\mathbf{x}|\theta) d\theta$ of the generated data.

@@

\begin{figure}
    \centering
    \includegraphics[width=0.48\textwidth]{figures/poisson.pdf}
    \caption{Discrete Poisson model with unknown scale.
             (Top left) Proposal distributions $p(\theta|\psi)$ after adversarial variational optimization. For both $\gamma=0$ and $\gamma=5$, the distributions correctly concentrate their density around
                        the true value $\log(\lambda^*)$. Penalizing the entropy of the proposal distribution ($\gamma=5$) results in a tighter density and therefore to better inference.
             (Top right) Model distributions $p(\mathbf{x}|\psi)$ after training.
             (Bottom) Empirical estimates of the variational upper bound $U_d$ as optimization progresses.
             }\label{fig:poisson}
\end{figure}

>>>>

Complex computer simulators are increasingly used across fields of science to
describe generative models tying parameters of an underlying theory to
experimental observations. Inference in this setup is often
difficult, as simulators rarely provide a way to directly evaluate the likelihood
function for a given observation. In this note, we develop a likelihood-free
inference algorithm for fitting a forward non-differentiable generative model to
(continuous or discrete) observed data. We adapt the adversarial training procedure of generative
adversarial networks by replacing the implicit generative network with a
domain-based scientific simulator, and solve the resulting non-differentiable
minimax problem by minimizing variational upper bounds of the adversarial objectives.
Effectively, the procedure results in learning an arbitrarily tight
proposal distribution over simulator parameters, such that the corresponding
marginal distribution of the generated data matches the observations.
\glnote{Mention experimental results.}
\glnote{Add 'so what?' conclusion.}

@@

In many fields of science such as particle physics, climatology or population
genetics, computer simulators are used to describe complex processes that tie
parameters of an underlying theory to high dimensional observations. In most
cases, these implicit generative models~\citep{2016arXiv161003483M} are
specified as procedural implementations of forward stochastic processes that
generate discrete or continuous data. Because it is usually computationally intractable, most simulators
do not provide a way to directly evaluate the likelihood function for a
given observation, thereby making inference difficult.
In addition, most scientific simulators are written using opaque low-level programming
languages, hence raising the bar for likelihood-free inference
algorithms relying e.g. on the simulator being a function with computable derivatives~\citep{2016arXiv160507826G}
or being a controllable probabilistic program~\citep{2016arXiv161009900L}.

In this note, we develop a likelihood-free inference algorithm for the point
estimation of the parameters of a forward non-differentiable
generative model. We adapt the adversarial
training procedure of generative adversarial
networks~\cite{goodfellow2014generative} by replacing the implicit generative
network with a domain-based scientific simulator, and solve the resulting
non-differentiable minimax problem by minimizing variational upper
bounds~\citep{2011arXiv1106.4487W,2012arXiv1212.4507S} of the adversarial
objectives. The procedure results in learning a proposal distribution over
simulator parameters, hence producing an arbitrarily tight family of models whose
joint collection of generated samples matches the observed data.

@@

Practically, the variational objectives \ref{eqn:vo-ud}-\ref{eqn:vo-ug}
have the effect of replacing the modeled data distribution of Eqn.~\ref{eqn:p_theta} with
a distribution parameterized in terms of $\psi$:
\begin{equation}\label{eqn:p_psi}
    \mathbf{x} \sim p(\mathbf{x}|\psi) \equiv \theta \sim q(\theta|\psi), \mathbf{z} \sim p(\mathbf{z}|\theta), \mathbf{x} = g(\mathbf{z}; \theta).
\end{equation}
Intuitively, this corresponds to a family of simulators, each configured
with randomly sampled parameters $\theta \sim q(\theta|\psi)$, whose joint collection
of generated samples is optimized with adversarial training to approach the real data distribution $p_r(\mathbf{x})$.
More formally, the learned model  therefore corresponds to the marginal distribution
of the generated data. That is,
\begin{equation}
    p(\mathbf{x}|\psi) = \int q(\theta|\psi) p(\mathbf{x}|\theta) d\theta.
\end{equation}

@@

\begin{figure}
    \centering
    \includegraphics[width=0.48\textwidth]{figures/poisson.pdf}
    \caption{Discrete Poisson model with unknown scale.
             (Top left) Proposal distributions $p(\theta|\psi)$ after adversarial variational optimization. For both $\gamma=0$ and $\gamma=5$, the distributions correctly concentrate their density around
                        the true value $\log(\lambda^*)$. Penalizing the entropy of the proposal distribution ($\gamma=5$) results in a tighter density and therefore to better inference.
             (Top right) Model distributions $p(\mathbf{x}|\psi)$ after training. This plot shows that the resulting parameterizations of the simulator closely reproduce the true distribution with, in this case, qualitatively
                        better results for the penalized model.
             (Bottom) Empirical estimates of the variational upper bound $U_d$ as optimization progresses.
             }\label{fig:poisson}
\end{figure}
